Sequential algorithms to split and merge ultra-high resolution 3D images

### 2017-08-11 paper.tex

\begin{abstract}
  Splitting and merging data is a requirement for most parallel or
  distributed processing operations. Naive algorithms, however,
  perform very poorly to split or merge 3D blocks from ultra-high
  resolution images, due to important seek times. We introduce and
  analyze sequential algorithms that leverage memory buffering, to
  address this issue. Cluster reads and Cluster writes access image
  chunks only once, but they have to seek in the reconstructed
  image. Multiple reads and Multiple writes minimize seeks in the
  reconstructed image, but they access image chunks multiple
  times. Evaluation on a 3850x3025x3500 brain image shows that our
  algorithms perform similarly to the optimal configuration (slice
  merging) provided that enough memory is available, which solves our
  initial problem to a large extent.  We conclude that splitting and
  merging large 3D images can be done efficiently without the need for
  complex data formats.
\end{abstract}

@@

\subsection{Buffered blocks: Cluster reads}

Cluster reads are the more direct extension of buffered slices to
blocks: they load multiple blocks in memory, concatenate them in a
buffer and write the buffer in the reconstructed image. Seeking is
reduced compared to naive block merging since contiguous parts of the
buffer will be written without seeking. A given
block is accessed only once during the whole merging process.

@@

\begin{figure}
\centering
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case1-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case2-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case3.pdf_tex}
\caption{Memory-load configurations in cluster reads, leading to
  different number of seeks. Red blocks need seeking before each of
  their rows ($d^2$ seeks). Blue blocks need seeking before each of
  their slices ($d$ seeks). Green blocks need only a single
  seek. Grey, dashed, transparent blocks represent the contiguous
  memory loads and are added for the sake of visualization.}
\label{fig:cluster-reads}
\end{figure}
Our cluster reads algorithm focuses on the three memory load
configurations represented in Figure~\ref{fig:cluster-reads}, that is,
the amount of memory $m'$ used by the algorithm is rounded down to the
closest number of complete blocks (case 1), of complete block rows (case
2) or of complete block slices (case 3). This is in general reasonable
since adding an incomplete row to a set of complete ones multiplies
the number of required seeks by $d$, as illustrated in
Figure~\ref{fig:avoided-configurations}-Left. In some cases though,
rounding $m$ down to $m'$ might increase the number of required
memory loads to a point that the overall number of seeks also
increases. Such cases are, however, slightly unusual and their
complete description requires extensive calculations involving modulo
arithmetic, which we felt were unwieldy to report here.
\begin{figure}
  \centering
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/incomplete-rows.pdf_tex}
\quad \quad \quad
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/overlap.pdf_tex}
\caption{Configurations that increase the number of seeks per
  memory-load and are thus deliberately avoided by cluster
  reads. Left: configuration with incomplete block rows (multiplies
  the number of seeks by $d$ compared to complete block rows). Right:
  configuration with block rows that overlap multiple block slices
  (multiplies the number of seeks by $2$ compared to non-overlapping
  configurations).}
\label{fig:avoided-configurations}
\end{figure}

Our algorithm also avoids configurations where the memory load
overlaps multiple block slices in case 2 or multiple block rows in
case 1, as such overlaps multiply the number of required seeks (see
Figure~\ref{fig:avoided-configurations}-Right).

Cluster reads are described in
Algorithm~\ref{algo:cluster-reads}. Function \texttt{switch} (line 3)
selects one of the three cases based on the amount of available memory
and the number of blocks. It returns \texttt{m'} and \texttt{case},
the identifier of the selected case. Function \texttt{check\_overlap}
(line 7) determines whether two blocks overlap multiple block slices
(case 2) or multiple block rows (case 1). For case 3 it always returns
false.  Function \texttt{sizeof} (line 8) returns the actual memory
used by its argument, including only its allocated segments in case
the argument is a buffer.
\begin{algorithm}[h]
  \caption{Buffered merging of blocks with cluster reads}
  \label{algo:cluster-reads}
  \begin{algorithmic}[1]
    \STATE sorted\_blocks = sort blocks by increasing (k,j,i)
    \STATE initialize buffer
    \STATE (m',case)=switch(m,n,R,b)
    \STATE old\_block = sorted\_blocks[0]
    \FOR{i = 0 ; i\textless n ; i+=1}
      \STATE block = sorted\_blocks[i]
      \STATE overlap = check\_overlap(block,old\_block,case)
      \IF{sizeof(buffer)+sizeof(block) $\geq$ m' \textbf{or} overlap=true}
      \STATE write buffer in reconstructed image
      \STATE clear buffer
      \STATE overlap = false
      \ENDIF
      \STATE read block and insert it in buffer
      \ENDFOR
  \end{algorithmic}
\end{algorithm}

@@

In the complexity analysis, we assume that $m'$ represents an integer
number $k$ of sub-rows (case 1, $k_1<\sqrt[3]{n}$), of complete rows
(case 2, $k_2<d$), of tile rows (case 3, $k_3<\sqrt[3]{n}$), of slices
(case 4, $k_4<d$) or of block slices (case 5, $k_5<\sqrt[3]{n}$), as
illustrated in Figure~\ref{fig:multiple-reads-cases}. In each of these
5 cases, we define $v_i$ as follows:
\begin{equation*}
  v_1=db \text{ ; }  v_2=Db \text{ ; } v_3=Ddb \text{ ; } v_4=D^2b \text{ ; } v_5=D^2db
\end{equation*}
so that we have:
\begin{equation*}
k_i=\floor*{\frac{m}{v_i}} \quad \mathrm{and} \quad m_i'=k_iv_i, \quad i \in \llbracket 1,5 \rrbracket
\end{equation*}

@@

Figure~\ref{fig:model-comparison} plots
Equations~\ref{eq:seeks-cluster-reads-1}
and~\ref{eq:seeks-multiple-reads} for different values of $n$. When
Cluster reads are in Case 1 or 2, they may outperform Multiple reads
only for large values of $n$. When Cluster reads are in case 3, they
are equivalent to Multiple reads: assuming that $Rb$ is an exact
multiple of $m$, Equations~\ref{eq:seeks-cluster-reads-1}
and~\ref{eq:seeks-multiple-reads} both boil down to $n+\frac{Rb}{m}$.
\todo{We could be a bit more rigorous and detail the cases where
  cluster reads of multiple reads will be better.}

\begin{figure}
  \includegraphics[width=0.45\columnwidth]{figures/model-big-brain-comparison.pdf}
  \includegraphics[width=0.45\columnwidth]{figures/model-big-brain-comparison-moreblocks.pdf}
  \caption{Number of seeks for Cluster reads vs Multiple reads, for
    D=3458 and b=2. Left: n=125; Right: n=64,000. Case 1 and Case 2
    denote Cluster read configurations.}
  % Gnuplot file: scripts/model/model.gnplt
  \label{fig:model-comparison}
\end{figure}

@@

% buffer implementation
The data buffer used in Cluster reads and Multiple reads is
implemented as a Python dictionary where the keys are offsets in the
reconstructed image and the values are NumPy arrays containing the
data starting at this offset. When the memory load is complete,
dictionary entries are written sequentially to the reconstructed
image. In Cluster reads, some seeking might be required between
writes. We implemented a defragmentation procedure for the dictionary
that merges contiguous dictionary entries in a single one, but we
abandoned it as it proved more time-consuming than going through all
the initial entries, due to the overhead of resizing NumPy arrays to
merge entries. In Multiple reads, dictionary entries are always
contiguous in the reconstructed image.  We tried to use a single NumPy
array as a buffer, but we finally abandoned it as inserting data at a
specific position in a NumPy array copies the data in memory, which
increases both the execution time and the peak memory consumption.

@@

@@

\subsection{Data}
We used the 3850x3025x3500 Big~Brain image split in 125
non-overlapping chunks of size 770x605x700 with 2 bytes per voxel
(total size, uncompressed is 75.92~GB). Big~Brain is a reference brain
based on the reconstruction of 7404 histological sections at nearly
cellular resolution of 20 micrometers~\cite{amunts2013bigbrain}. It is
a freely, publicly available tool with numerous applications in
neurosciences and neurosurgery.

We used the blocks of the 2015 Big~Brain release with 40-micrometer
isotropic resolution available at
\url{ftp://bigbrain.loris.ca/BigBrainRelease.2015/3D\_Blocks/40um}. We
converted them to NiFTI 1.0 using Nibabel and left them uncompressed.
We generated NiFTI slices of size 3850x3025x28 from the reconstructed
image, using our naive split algorithm. Slices are available at
\todo{\url{x}}.

@@

We used a Dell Precision Tower 3620 workstation with CentOS Linux
release 7.3.1611, 32~GB of RAM and two disks. (1) a Hard disk drive
(HDD): HGST Travelstar 7K1000, 7200~rpm, 931GiB (1TB), firmware
version JB0OA3W0; (2) a Solid-state drive (SSD): SanDisk X400 2.5,
238GiB (256GB), firmware version X4130012.  Both drives used 512-byte
logical sectors, 4096-byte physical sectors, SATA \textgreater 3.1 (6.0 Gb/s) and
were accessed through the XFS file system v4.5.0. We used
\texttt{iotop} (\url{http://guichaz.free.fr/iotop}) to monitor I/Os on
the workstation and make sure that no other process was compromising
our measures.
%smartctl --xall /dev/sda

@@

We used Git tag 0.1 of our \texttt{sam} library. Our experiment
scripts are available at
\url{https://github.com/big-data-lab-team/paper-sequential-split-merge/blob/master/scripts/experiment}. We
used them to split and merge using Buffered slices, Cluster reads and
Multiple reads, with 3~GB, 6~GB, 9~GB, 12~GB and 16~GB of
memory. Table~\ref{table:configs} shows the configurations of Cluster
reads and Multiple reads for each memory value, according to
Equations~\ref{eq:seeks-cluster-reads-1}
and~\ref{eq:seeks-multiple-reads}. For instance, Cluster reads are in
case 1 for 3~GB. We also did 0~GB of memory for Buffered slices and
Cluster reads, which triggered naive slice and block merging. We did 5
repetitions for each memory value. Memory values were shuffled in each
repetition, to correct for potential biases coming from ordering, such
as caching effects. To improve reproducibility, we dropped the kernel
page, dentry and inode caches before each run (\texttt{echo 3 | sudo
  tee /proc/sys/vm/drop\_caches}). We measured the cumulative read,
write and seek time in each run, as well as the overhead time defined
as the total time minus the sum of all other times.
\begin{table}
  \centering
\begin{footnotesize}
\begin{tabular}{|c|ccccc|}
  \hline
                 & 3~GB & 6~GB & 9~GB & 12~GB & 16~GB\\
  \hline
  Cluster  reads &  1 &  2 &  2 &  2 &  3 \\
  Multiple reads &  4 &  4 &  4 &  4 &  5\\
  \hline
\end{tabular}
\end{footnotesize}
\caption{Algorithm configurations by memory values.}
\label{table:configs}
\end{table}

@@

@@

The number of seeks is reported in Figure~\ref{fig:number-of-seeks},
for all algorithms and the corresponding models
(Equations~\ref{eq:naive-blocks}
to~\ref{eq:seeks-multiple-reads}). Note the logarithmic y scale. Error
bars are not reported as numbers of seeks were constant across all
repetitions. The mean model error is \todo{x\%}, explained by the fact
that the model assumes cubic blocks while we used non-cubic ones in
the experiment. Overall, the model is validated as it correctly
explains the observations.  
\begin{figure}[h]
  \centering
  \includegraphics[width=\columnwidth]{figures/number-of-seeks.pdf}
  \hfill
  \caption{Number of seeks for all algorithms. y scale is logarithmic. Each algorithm is
    represented with a different color. Dark color is experimental
    value; bright color is model.}
  % Gnuplot file: scripts/number_of_seeks.gnuplot
\label{fig:number-of-seeks}
\end{figure}

As expected, the difference between naive blocks and naive slices is
tremendous, in the order of 50 million seeks. For 3~GB, the number of
seeks in Cluster reads is 4 orders of magnitude higher than for
Multiple reads, and 5 orders of magnitude higher than naive
slices. This huge difference comes from the fact that at 3~GB Cluster
reads are in case 1. For 6~GB, 9~GB and 12~GB, Cluster reads are in
case 2 and the difference with Multiple reads and Buffered slices
reduces. At 16~GB, all algorithms perform the same.

@@

\todo{ The measured seek times correlated with
the measured merging times ($\rho$=x), which suggest that the
observed time reductions were indeed caused by reduced seek times.}

% Slices and blocks
Figure~\ref{fig:merge-time} shows the merge time for all algorithms by
memory values. Naive blocks take about 5 times longer than naive
slices to merge the data, which quantifies the effect of the problem
targeted by our algorithms. This effect appears both on HDD and SSD,
which shows that SSDs might not be the only solution to reduce seek
time. In the remainder we use naive blocks and naive slices and
references to evaluate our algorithms.

@@

% Cluster reads
Cluster reads provide important speed ups compared to naive blocks,
both on HDD and on SSD. On SSD, they are up to 5 times faster than
naive blocks at 16~GB, and 3 on average \todo{compute the actual
  values}. \todo{On HDD, \ldots} Surprisingly, they perform
substantially faster than naive blocks even at 3~GB, while in case
1. This may be explained by the fact that the seeks required to write
incomplete block rows to the reconstructed image are shorter than the
ones for naive blocks.

% Multiple reads
Multiple reads are even faster than Cluster reads. On SSD, they are 5
times faster than naive blocks on average \todo{compute the actual
  values} ( \todo{On HDD, \ldots}). At 12~GB, the total merging time
is only 1.2 times slower than naive slices, which solves our initial
problem to a large extent. Even at 3~GB the speed-up is considerable
compared to naive blocks.

@@

\subsubsection{Merge time breakdown}
Figure~\ref{fig:breakdowns-ssd} shows how the total merge time breaks
down to read, write, seek and overhead time for our algorithms. Naive
blocks and naive slices are shown as references. The huge difference
between naive blocks and naive slices is coming from both the seek
time and the write time, which suggests that seeking degrades the
write rate importantly in addition to introducing extra delays.
Cluster reads reduce the seek time and thus the read time very
substantially. Multiple reads almost anihilate the seek time and bring
the read time to a value comparable to naive slices.  The same
behavior is observed on HDD and on SSD, although the effect of seeking
is slightly lower on SSD, as expected. Read times are consistently and substantially lower than write times\todo{explain why}. The overhead time is small for
both Cluster reads and Multiple reads, although it is not negligible
for Multiple reads on HDD.
\begin{figure}
  \begin{subfigure}[b]{\columnwidth}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-buff-slices/buff-slices-breakdown-hdd.pdf}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-buff-slices/buff-slices-breakdown-ssd.pdf}
    \caption{Buffered slices}
  \end{subfigure}
  \begin{subfigure}[b]{\columnwidth}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-creads/creads-breakdown-hdd.pdf}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-creads/creads-breakdown-ssd.pdf}
    \caption{Cluster reads}
  \end{subfigure}
  \begin{subfigure}[b]{\columnwidth}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-mreads/mreads-breakdown-hdd.pdf}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-mreads/mreads-breakdown-ssd.pdf}
    \caption{Multiple reads}
  \end{subfigure}
  \caption{Breakdown of total merge times. Left column: HDD. Right column: SSD.}
  \label{fig:breakdowns-ssd}
  % Gnuplot file: scripts/experiment/compare_breakdown_errbar.gnuplot
\end{figure}

@@

\todo{define cluster writes and multiple writes somewhere.}

The number of seeks for Cluster writes and Multiple writes were
identical to the ones in Cluster reads and Multiple reads. \todo{Check
  that.} The total split time by algorithm is shown in
Figure~\ref{fig:split-time} and the breakdown by read, write, seek and
overhead time is in Figure~\ref{fig:breakdowns-ssd-split}. \todo{comment on the figures when results are available.}

@@

\begin{figure}
  \begin{subfigure}[b]{\columnwidth}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-buff-slices/buff-slices-breakdown-hdd-split.pdf}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-buff-slices/buff-slices-breakdown-ssd-split.pdf}
    \caption{Buffered slices}
  \end{subfigure}
  \begin{subfigure}[b]{\columnwidth}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-cwrite/cwrites-breakdown-hdd.pdf}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-cwrites/cwrites-breakdown-ssd.pdf}
    \caption{Cluster writes}
  \end{subfigure}
  \begin{subfigure}[b]{\columnwidth}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-mwrites/mwrites-breakdown-hdd.pdf}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-mwrites/mwrites-breakdown-ssd.pdf}
    \caption{Multiple writes}
  \end{subfigure}
  \caption{Breakdown of total split times. Left column: HDD. Right column: SSD.}
  \label{fig:breakdowns-ssd-split}
  % Gnuplot file: scripts/experiment/compare_breakdown_errbar.gnuplot
\end{figure}

@@

% Parallel algorithms
In a real system, high-resolution images are likely to be processed on
computing clusters, for instance using software from the Hadoop
project, in particular the Hadoop Distributed File
System~\cite{shvachko2010hadoop}. In this context parallel split-and-merge algorithms
will be required, since the various blocks of a large image would be
uploaded to different disks concurrently. In the same vein,
``re-spliting'' algorithms would have to be devised in case a split
image need to be split in a different geometry. Cluster reads and
multiple reads could be used as starting points for such algorithms.

>>>>

\begin{abstract}
  Splitting and merging data is a requirement for most parallel or
  distributed processing operations. Naive algorithms, however,
  perform very poorly to split or merge 3D blocks from ultra-high
  resolution images, due to important seek times. We introduce and
  analyze sequential algorithms that leverage memory buffering, to
  address this issue. Clustered reads and Clustered writes access image
  chunks only once, but they have to seek in the reconstructed
  image. Multiple reads and Multiple writes minimize seeks in the
  reconstructed image, but they access image chunks multiple
  times. Evaluation on a 3850x3025x3500 brain image shows that our
  algorithms perform similarly to the optimal configuration (slice
  merging) provided that enough memory is available, which solves our
  initial problem to a large extent.  We conclude that splitting and
  merging large 3D images can be done efficiently without the need for
  complex data formats.
\end{abstract}

@@

\subsection{Buffered blocks: Clustered reads}

Clustered reads are the more direct extension of buffered slices to
blocks: they load multiple blocks in memory, concatenate them in a
buffer and write the buffer in the reconstructed image. Seeking is
reduced compared to naive block merging since contiguous parts of the
buffer will be written without seeking. A given
block is accessed only once during the whole merging process.

@@

\begin{figure}
\centering
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case1-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case2-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case3.pdf_tex}
\caption{Memory-load configurations in cluster reads, leading to
  different number of seeks. Red blocks need seeking before each of
  their rows ($d^2$ seeks). Blue blocks need seeking before each of
  their slices ($d$ seeks). Green blocks need only a single
  seek. Grey, dashed, transparent blocks represent the contiguous
  memory loads and are added for the sake of visualization.}
\label{fig:cluster-reads}
\end{figure}
Our cluster reads algorithm focuses on the three memory load
configurations represented in Figure~\ref{fig:cluster-reads}, that is,
the amount of memory $m'$ used by the algorithm is rounded down to the
closest number of complete blocks (Case 1), of complete block rows (Case
2) or of complete block slices (Case 3). This is in general reasonable
since adding an incomplete row to a set of complete ones multiplies
the number of required seeks by $d$, as illustrated in
Figure~\ref{fig:avoided-configurations}-Left. In some cases though,
rounding $m$ down to $m'$ might increase the number of required
memory loads to a point that the overall number of seeks also
increases. Such cases are, however, slightly unusual and their
complete description requires extensive calculations involving modulo
arithmetic, which we felt were unwieldy to report here.
\begin{figure}
  \centering
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/incomplete-rows.pdf_tex}
\quad \quad \quad
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/overlap.pdf_tex}
\caption{Configurations that increase the number of seeks per
  memory-load and are thus deliberately avoided by cluster
  reads. Left: configuration with incomplete block rows (multiplies
  the number of seeks by $d$ compared to complete block rows). Right:
  configuration with block rows that overlap multiple block slices
  (multiplies the number of seeks by $2$ compared to non-overlapping
  configurations).}
\label{fig:avoided-configurations}
\end{figure}

Our algorithm also avoids configurations where the memory load
overlaps multiple block slices in Case 2 or multiple block rows in
Case 1, as such overlaps multiply the number of required seeks (see
Figure~\ref{fig:avoided-configurations}-Right).

Clustered reads are described in
Algorithm~\ref{algo:cluster-reads}. Function \texttt{switch} (line 3)
selects one of the three cases based on the amount of available memory
and the number of blocks. It returns \texttt{m'} and \texttt{case},
the identifier of the selected case. Function \texttt{check\_overlap}
(line 7) determines whether two blocks overlap multiple block slices
(case 2) or multiple block rows (Case 1). For case 3 it always returns
false.  Function \texttt{sizeof} (line 8) returns the actual memory
used by its argument, including only its allocated segments in case
the argument is a buffer.
\begin{algorithm}[h]
  \caption{Buffered merging of blocks with cluster reads}
  \label{algo:cluster-reads}
  \begin{algorithmic}[1]
    \STATE sorted\_blocks = sort blocks by increasing (k,j,i)
    \STATE initialize buffer
    \STATE (m',case)=switch(m,n,R,b)
    \STATE old\_block = sorted\_blocks[0]
    \FOR{i = 0 ; i\textless n ; i+=1}
      \STATE block = sorted\_blocks[i]
      \STATE overlap = check\_overlap(block,old\_block,case)
      \IF{sizeof(buffer)+sizeof(block) $\geq$ m' \textbf{or} overlap=true}
      \STATE write buffer in reconstructed image
      \STATE clear buffer
      \STATE overlap = false
      \ENDIF
      \STATE read block and insert it in buffer
      \ENDFOR
  \end{algorithmic}
\end{algorithm}

@@

In the complexity analysis, we assume that $m'$ represents an integer
number $k$ of sub-rows (Case 1, $k_1<\sqrt[3]{n}$), of complete rows
(Case 2, $k_2<d$), of tile rows (Case 3, $k_3<\sqrt[3]{n}$), of slices
(Case 4, $k_4<d$) or of block slices (Case 5, $k_5<\sqrt[3]{n}$), as
illustrated in Figure~\ref{fig:multiple-reads-cases}. In each of these
5 cases, we define $v_i$ as follows:
\begin{equation*}
  v_1=db \text{ ; }  v_2=Db \text{ ; } v_3=Ddb \text{ ; } v_4=D^2b \text{ ; } v_5=D^2db
\end{equation*}
so that we have:
\begin{equation*}
k_i=\floor*{\frac{m}{v_i}} \quad \mathrm{and} \quad m_i'=k_iv_i, \quad i \in \llbracket 1,5 \rrbracket
\end{equation*}

@@

Figure~\ref{fig:model-comparison} plots
Equations~\ref{eq:seeks-cluster-reads-1}
and~\ref{eq:seeks-multiple-reads} for different values of $n$. When
Clustered reads are in Case 1 or 2, they may outperform Multiple reads
only for large values of $n$. When Clustered reads are in case 3, they
are equivalent to Multiple reads: assuming that $Rb$ is an exact
multiple of $m$, Equations~\ref{eq:seeks-cluster-reads-1}
and~\ref{eq:seeks-multiple-reads} both boil down to $n+\frac{Rb}{m}$.
\todo{We could be a bit more rigorous and detail the cases where
  cluster reads of multiple reads will be better.}

\begin{figure}
  \includegraphics[width=0.45\columnwidth]{figures/model-big-brain-comparison.pdf}
  \includegraphics[width=0.45\columnwidth]{figures/model-big-brain-comparison-moreblocks.pdf}
  \caption{Number of seeks for Clustered reads vs Multiple reads, for
    D=3458 and b=2. Left: n=125; Right: n=64,000. Case 1 and Case 2
    denote Clustered read configurations.}
  % Gnuplot file: scripts/model/model.gnplt
  \label{fig:model-comparison}
\end{figure}

@@

% buffer implementation
The data buffer used in Clustered reads and Multiple reads is
implemented as a Python dictionary where the keys are offsets in the
reconstructed image and the values are NumPy arrays containing the
data starting at this offset. When the memory load is complete,
dictionary entries are written sequentially to the reconstructed
image. In Clustered reads, some seeking might be required between
writes. We implemented a defragmentation procedure for the dictionary
that merges contiguous dictionary entries in a single one, but we
abandoned it as it proved more time-consuming than going through all
the initial entries, due to the overhead of resizing NumPy arrays to
merge entries. In Multiple reads, dictionary entries are always
contiguous in the reconstructed image.  We tried to use a single NumPy
array as a buffer, but we finally abandoned it as inserting data at a
specific position in a NumPy array copies the data in memory, which
increases both the execution time and the peak memory consumption.

@@

We also implemented the splitting algorithms corresponding to
Clustered reads and Multiple reads, called Clustered writes and
Multiple writes.

@@

\subsection{Data}
We used the 3850x3025x3500 Big~Brain image split in 125
non-overlapping chunks of size 770x605x700 with 2 bytes per voxel
(total size uncompressed is 75.92~GB). Big~Brain is a reference brain
based on the reconstruction of 7404 histological sections at nearly
cellular resolution of 20 micrometers~\cite{amunts2013bigbrain}. It is
a freely, publicly available tool with numerous applications in
neurosciences and neurosurgery.

We used the blocks of the 2015 Big~Brain release with 40-micrometer
isotropic resolution available at
\url{ftp://bigbrain.loris.ca/BigBrainRelease.2015/3D\_Blocks/40um}. We
converted them to NiFTI 1.0 using Nibabel and left them uncompressed.
We generated NiFTI slices of size 3850x3025x28 from the reconstructed
image, using our naive split algorithm. Slices are available at
\todo{Upload them (compressed) to figshare.}.

@@

We used a Dell Precision Tower 3620 workstation with CentOS Linux
release 7.3.1611, 32~GB of RAM and two disks: (1) a Hard Disk Drive
(HDD): HGST Travelstar 7K1000, 7200~rpm, 931GiB (1TB), firmware
version JB0OA3W0; (2) a Solid-state drive (SSD): SanDisk X400 2.5,
238GiB (256GB), firmware version X4130012.  Both drives used 512-byte
logical sectors, 4096-byte physical sectors, SATA \textgreater 3.1 (6.0 Gb/s) and
were accessed through the XFS file system v4.5.0. We used
\texttt{iotop} (\url{http://guichaz.free.fr/iotop}) to monitor I/Os on
the workstation and make sure that no other process was compromising
our measures.
%smartctl --xall /dev/sda

@@

We used Git tag 0.1 of our \texttt{sam} library \todo{tag it}. Our
experiment scripts are available at
\url{https://github.com/big-data-lab-team/paper-sequential-split-merge/blob/master/scripts/experiment}. We
used them to split and merge using Buffered slices, Clustered reads and
Multiple reads, with 3~GB, 6~GB, 9~GB, 12~GB and 16~GB of
memory. Table~\ref{table:configs} shows the configurations of Clustered
reads and Multiple reads for each memory value, according to
Equations~\ref{eq:seeks-cluster-reads-1}
and~\ref{eq:seeks-multiple-reads}. For instance, at 3~GB, Clustered reads were in
Case 1. We also did a run with 0~GB of memory for Buffered slices and
Clustered reads, which triggered naive slice and block merging. We did 5
repetitions for each memory value. Memory values were shuffled in each
repetition, to avoid potential ordering biases such
as caching effects. To ensure equal conditions, we dropped the kernel
page, dentry and inode caches before each run (\texttt{echo 3 | sudo
  tee /proc/sys/vm/drop\_caches}). We measured the cumulative read,
write and seek time in each run, as well as the overhead time defined
as the total time minus the sum of all other times.
\begin{table}
  \centering
\begin{footnotesize}
\begin{tabular}{|c|ccccc|}
  \hline
                 & 3~GB & 6~GB & 9~GB & 12~GB & 16~GB\\
  \hline
  Clustered  reads &  1 &  2 &  2 &  2 &  3 \\
  Multiple reads &  4 &  4 &  4 &  4 &  5\\
  \hline
\end{tabular}
\end{footnotesize}
\caption{Algorithm configurations by memory values.}
\label{table:configs}
\end{table}

@@

All the experimental data and scripts used to generate the figures in
this Section are available at
\url{https://github.com/big-data-lab-team/paper-sequential-split-merge}
under GPLv3 license.

@@

The number of seeks is reported in Figure~\ref{fig:number-of-seeks},
for all algorithms and the corresponding models
(Equations~\ref{eq:naive-blocks}
to~\ref{eq:seeks-multiple-reads}). Note the logarithmic y scale. Error
bars are not reported as numbers of seeks were constant across all
repetitions. The average relative model errors are 12.7\% (Naive
blocks), 0\% (Naive slices), 3.3\% (Clustered reads), 26.8\% (Multiple
reads) and 0.9\% (Buffered slices), presumably explained by the fact
that the model assumes cubic blocks while we used non-cubic ones in
the experiment. For Multiple reads, our complexity analysis also
assumed that one of the 5 Cases in
Figure~\ref{fig:multiple-reads-cases} was used while they are usually
blended in practice. Overall, the model correctly explains the
observations.
\begin{figure}[h]
  \centering
  \includegraphics[width=\columnwidth]{figures/number-of-seeks.pdf}
  \hfill
  \caption{Number of seeks for all algorithms. y scale is logarithmic. Each algorithm is
    represented with a different color. Dark color is experimental
    value; bright color is model.}
  % Gnuplot file: scripts/number_of_seeks.gnuplot
\label{fig:number-of-seeks}
\end{figure}

As expected, the difference between naive blocks and naive slices is
tremendous, in the order of 50 million seeks. For 3~GB, the number of
seeks in Clustered reads is 4 orders of magnitude higher than for
Multiple reads, and 5 orders of magnitude higher than naive
slices. This huge difference comes from the fact that at 3~GB Clustered
reads are in Case 1. For 6~GB, 9~GB and 12~GB, Clustered reads are in
case 2 and the difference with Multiple reads and Buffered slices
reduces. At 16~GB, all algorithms perform the same.

@@

% Slices and blocks
Figure~\ref{fig:merge-time} shows the merge time for all algorithms by
memory values. Naive blocks are 9.5 times slower than naive slices on
HDD (6.7 times on SSD), which quantifies the effect of the targeted
problem. It also shows that SSDs might not be the only solution to
reduce the impact of seek time. In the remainder we use naive blocks
and naive slices as references to evaluate our algorithms.

@@

% Clustered reads
Clustered reads provide important speed ups compared to naive blocks,
both on HDD and on SSD. On HDD, they are up to 9.7 times faster than
naive blocks at 16~GB and 6.8 times faster on average. On SSD, they
are up to 5.1 times faster than naive blocks at 16~GB, and 3.1 times
faster on average.  Surprisingly, they perform substantially faster
than naive blocks even at 3~GB, while in Case 1. This may be explained
by the fact that the seeks required to write incomplete block rows to
the reconstructed image are shorter than the ones for naive blocks.

% Multiple reads
Multiple reads are even faster than Clustered reads on this dataset. On
HDD, they are 8.4 times faster than naive blocks on average (5.3 times
on SSD). At 16~GB, they are 9.0 times faster than naive blocks on
average (5.8 times faster on SSD).

@@

\subsubsection{Merge time breakdown}
Figure~\ref{fig:breakdowns-ssd} shows how the total merge time breaks
down to read, write, seek and overhead time for our algorithms. Naive
blocks and naive slices are shown as references. The huge difference
between naive blocks and naive slices is coming from both the seek
time and the write time, which suggests that seeking degrades the
write rate importantly in addition to introducing extra delays.
Clustered reads reduce the seek time and thus the read time very
substantially. Multiple reads almost anihilate the seek time and bring
the read time to a value comparable to naive slices.  The same
behavior is observed on HDD and on SSD, although the effect of seeking
is slightly smaller on SSD, as expected. Read times are consistently
and substantially lower than write times. This may be a result of
reading data using Python's NumPy package, which is more efficient
than using native Python - as is the case with our writes. The
overhead time is small for both Clustered reads and Multiple reads.
\begin{figure}
  \begin{subfigure}[b]{\columnwidth}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-buff-slices/buff-slices-breakdown-hdd.pdf}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-buff-slices/buff-slices-breakdown-ssd.pdf}
    \caption{Buffered slices}
  \end{subfigure}
  \begin{subfigure}[b]{\columnwidth}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-creads/creads-breakdown-hdd.pdf}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-creads/creads-breakdown-ssd.pdf}
    \caption{Clustered reads}
  \end{subfigure}
  \begin{subfigure}[b]{\columnwidth}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-mreads/mreads-breakdown-hdd.pdf}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-mreads/mreads-breakdown-ssd.pdf}
    \caption{Multiple reads}
  \end{subfigure}
  \caption{Breakdown of total merge times. Left column: HDD. Right column: SSD.}
  \label{fig:breakdowns-ssd}
  % Gnuplot file: scripts/experiment/compare_breakdown_errbar.gnuplot
\end{figure}

@@

The number of seeks for Clustered writes and Multiple writes were
identical to the ones in Clustered reads and Multiple reads. \todo{Check
  that.} The total split time by algorithm is shown in
Figure~\ref{fig:split-time} and the breakdown by read, write, seek and
overhead time is in Figure~\ref{fig:breakdowns-ssd-split}. \todo{comment on the figures when results are available.}

@@

\begin{figure}
  \begin{subfigure}[b]{\columnwidth}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-buff-slices/buff-slices-breakdown-hdd-split.pdf}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-buff-slices/buff-slices-breakdown-ssd-split.pdf}
    \caption{Buffered slices}
  \end{subfigure}
  \begin{subfigure}[b]{\columnwidth}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-cwrite/cwrites-breakdown-hdd.pdf}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-cwrites/cwrites-breakdown-ssd.pdf}
    \caption{Clustered writes}
  \end{subfigure}
  \begin{subfigure}[b]{\columnwidth}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-mwrites/mwrites-breakdown-hdd.pdf}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-mwrites/mwrites-breakdown-ssd.pdf}
    \caption{Multiple writes}
  \end{subfigure}
  \caption{Breakdown of total split times. Left column: HDD. Right column: SSD.}
  \label{fig:breakdowns-ssd-split}
  % Gnuplot file: scripts/experiment/compare_breakdown_errbar.gnuplot
\end{figure}

@@
